Prediction of Protein-Protein Interactions on the Human and Rice Interactomes

### 2020-12-09 nicolas/PPIPL3/ReportV6.tex
files were missing in git

%% LyX 2.3.4.2 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[english]{article}
\usepackage[T1]{fontenc}
\usepackage[latin9]{inputenc}
\usepackage{verbatim}
\usepackage{float}
\usepackage{graphicx}

@@

\usepackage{babel}
\begin{document}
\title{Prediction of Protein-Protein Interactions on the Human and Rice Interactome }
\author{Nicolás López-Rozo\thanks{Pontificia Universidad Javeriana Cali, ORCID: 0000-0002-8060-6891\protect \\
email: nicolaslopez@javerianacali.edu.co}, Jorge Finke\thanks{Pontificia Universidad Javeriana Cali, ORCID: 0000-0003-2393-6066},
Camilo Rocha\thanks{Pontificia Universidad Javeriana Cali, ORCID: 0000-0003-4356-7704}}
\maketitle
\begin{abstract}

Efforts to predict unmapped protein-protein interactions (PPI) through
network analysis suggest that the higher the number of paths of length
3 (L3) between two proteins, the higher the probability of interaction.
This paper extends previous work on predicting protein interactions
based on the L3 principle by taking into account the learning of node
features of the PPI network. In particular, we train an \texttt{XGBoost}
model using handcrafted features (including L3), as well as edge embeddings
from \texttt{Node2Vec}. Our main result shows that while L3 is an
important feature for predicting links, best performance is achieved
when combined with edge embeddings. The proposed approach is evaluated
for the human and rice interactomes and pretends to exploit the intrinsic
network structure. Results for both organisms show higher AUC values
for the combination of automated and handcrafted features than for
each of its components alone. The handcrafted features have more importance
for the prediction model in the human interactome than in rice. The
developed framework can be easily applied to interactomes for other
organisms or even to networks outside of the biological context.
\end{abstract}
\begin{comment}
need include a feature that compares the dimension of the embedding
is useful for prediction with L3 to show that they independent measures
- we do not want to be learning L3 through node2vec
\end{comment}

@@

\begin{comment}
Need to add other relevant references
\end{comment}

Proteins are key actors of biological processes inside cells. As they
carry out a variety of tasks, proteins do not work as single agents
but rather as a part of dynamic networks of protein-protein interactions
(PPI) \cite{Lin2017}. PPI networks play an important role in a variety
of interdependent mechanisms, including signal transduction, homeostasis
control, stress responses, plant defense and organ formation. They
are also involved a variety of physiological and developmental processes,
including protein phosphorylation, transcriptional co-factor recruitment
and transporter activation \cite{Zhang_2010_PPI}.

Different methods for predicting interactions in PPI networks have
been proposed in recent years \cite{Chen2019PPI,Chang2016PPI,Kotlyar2015PPI}.
Kovacs et al. (2019) introduce an approach which relies on counting
the number of paths of length 3 (L3) to predict interactions among
proteins for different organisms. The proposed approach outperforms
previous methods in PPI networks of yeast (\emph{S. cerevisiae}),
Arabidopsis (\emph{A. thaliana}), worm (\emph{C. elegans}), fly (\emph{D.
melanogaster}), fission yeast (\emph{S. pombe}), mouse (\emph{M. musculus})
and the human interactome \cite{Kovacs2019}.

A common way to uncover (or validate) PPIs is the \emph{Yeast-Two-Hybrid}
technique (also known as \emph{two-hybrid screening} or \emph{Y2H}),
which is based on (i) the expression of a specific reporter gene that
activates by the binding of a DNA-binding Domain (DB), and (ii) an
Activation Domain (AD) of a Transcription Factor that binds to an
Upstream Activation Sequence (UAS). For the Y2H technique, a protein
is fused to the DB domain (known as \emph{bait}) and another protein
to the AD (known as \emph{prey}). If the proteins do not interact,
then the reporter gene is not expressed. Otherwise, the reporter gene
expression is activated by the activation domain.

\begin{figure}[h]
\caption{Yeast-2-Hybrid Technique}

@@

Based on the outcome of numerous Y2H experiments, PPI networks which
capture known interactions for each protein can be constructed. Several
algorithms are proposed over these networks to predict unknown interactions.
The focus of this study is to evaluate different methods for predicting
PPIs using the network of known interactions. These interactions are
represented as an undirected graph. A common way of predicting interactions
is usually based on social networks analysis, more specifically on
the triadic closure principle (TCP). TCP states that the higher the
number of common neighboring nodes between two nodes, the higher the
probability that they interact \cite{Goldberg2003SmallWorld}. Previous
studies show that the mentioned approach fails because it does not
consider the structural and chemical properties of the proteins \cite{Kovacs2019,Cannistraci2013Networks}.

@@

For achieving the described results, human and rice PPI networks are
compared using state-of-the-art methods, as well as the proposed ones
(A2, A3, L3). In the case of the human network, two different versions
of the human interactome (\emph{HI-II-14}, \emph{HI-TESTED})\cite{Rolland2014Human}
are used for training and another interactome version is used for
validation (\emph{HI-III})\cite{Kovacs2019}. For the rice interactome,
a portion of the known interactions removed and then predicted with
different techniques based on the known interactions. The main contributions
of this paper are i) a general framework for link prediction in non-directed
networks and ii) two applications of this framework to biological
networks with a structural insight

\begin{comment}
JF: read throughout the document but edited up here. I prefer to read
the results before editing the materials and methods section. 
\end{comment}

@@

Importance values for the A3 method stand out less from the node embedding
information than in A2 and L3. In this case, the importance gain of
the handcrafted features in each model was not as high in rice as
in the human interactome.

>>>>

%% LyX 2.3.4.2 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[english]{article}
\usepackage[T1]{fontenc}
\usepackage[latin9]{inputenc}
\usepackage{verbatim}
\usepackage{float}
\usepackage{textcomp}
\usepackage{graphicx}

@@

\usepackage{babel}
\begin{document}
\title{Prediction of Protein-Protein Interactions on the Human and Rice Interactome }
\author{Nicolás López-Rozo\thanks{Pontificia Universidad Javeriana Cali, ORCID: 0000-0002-8060-6891\protect \\
\protect\hphantom{ema.}email: nicolaslopez@javerianacali.edu.co}, Jorge Finke\thanks{Pontificia Universidad Javeriana Cali, ORCID: 0000-0003-2393-6066},
Camilo Rocha\thanks{Pontificia Universidad Javeriana Cali, ORCID: 0000-0003-4356-7704}}
\maketitle
\begin{abstract}

Recent developments in network-based prediction of unmapped protein-protein
interactions (PPIs) reveal that two proteins are more likely to interact,
the higher the number of paths of length 3 between them, normalized
by the geometric average of their interactions. This paper extends
previous work on predicting binary interactions by taking into account
the learning of features (embeddings) of the PPI network. In particular,
we implement a gradient boosted decision tree model (\texttt{XGBoost})
using handcrafted features (including the normalized measure of paths
of length 3, L3) and embeddings from an algorithm that generates a
low-dimensional representation of nodes (\texttt{Node2Vec}). Our main
result shows that while L3 remains an important feature for predicting
PPIs, a better performance is achieved when L3 is combined with embedding
features. The proposed approach is validated for the human and rice
interactomes. For both cases, higher AUC values are obtained when
both types of features are used to train the model. The developed
framework can also be applied to interactomes of other organisms for
which a PPI network become recently available.
\end{abstract}

@@

Proteins are key actors of biological processes inside cells. Rather
than carrying out tasks as single agents, they are part of dynamic
networks of protein-protein interactions (PPI) \cite{Lin2017}. Such
networks underlie a variety of interdependent mechanisms, including
signal transduction, homeostasis control and stress responses. Furthermore,
PPI networks play an important role in physiological and developmental
processes such as protein phosphorylation, transcriptional co-factor
recruitment and transporter activation \cite{Zhang_2010_PPI}.

A common way to create PPI networks or validate particular protein-protein
interactions is the \emph{Yeast-Two-Hybrid} technique (also known
as \emph{two-hybrid screening} or \emph{Y2H}). Figure \ref{Y2H}A
illustrates the biological basis of Y2H: the expression of a specific
reporter gene that activates by the binding of a DNA-binding Domain
(DB) and an Activation Domain (AD) of a Transcription Factor that
binds to an Upstream Activation Sequence (UAS). To evaluate an interaction,
the Y2H approach fuses a protein to the DB domain (known as \emph{bait})
and another protein to the AD (known as \emph{prey}). If the proteins
interact, then the reporter gene expression is activated by the AD
(Fig. \ref{Y2H}B). Otherwise, if proteins fail to interact, then
the reporter gene is not expressed (Fig. \ref{Y2H}C).

\begin{figure}[h]
\caption{\label{Y2H}The Yeast-2-Hybrid technique offers an experimental approach
for constructing PPI networks}

@@

Based on the outcome of numerous Y2H experiments, undirected networks
of interactions between proteins can be constructed. However, relying
only on experimental validation of PPIs is detrimental to research
due to its hindrance in costs, accuracy, required manpower and time
\cite{Laraia2015PPI,Macalino2018PPI}. Besides, the number of missing
interactions for pairs of proteins on many organisms leverage the
usage of computational methods for prediction of such interactions.
\begin{comment}
The focus of this study is to evaluate different methods for predicting
PPIs using the network of known interactions. 
\end{comment}

A common way of predicting interactions is usually based on social
networks analysis, more specifically on the triadic closure principle
(TCP). TCP states that the higher the number of common neighboring
nodes between two nodes, the higher the probability that they interact
\cite{Goldberg2003SmallWorld}. TCP can be addressed mathematically
by counting the number of shared neighbors of a pair of nodes, also
known as the Common Neighbors algorithm. By raising the adjacency
matrix (A) of the network to the second power (A\texttwosuperior ),
TCP can be considered for further analyses. However, previous studies
show that the mentioned approach usually fails because it does not
consider the structural and chemical properties of the proteins \cite{Cannistraci2013Networks,Kovacs2019}.

A variety of methods for predicting interactions in PPI networks have
been proposed in recent years \cite{Chang2016PPI,Chen2019PPI,Kotlyar2015PPI}.
Kovacs et al (2019) introduce a network-based approach which predicts
the interaction between two proteins based on the number of paths
of length 3, normalized by the geometric average of their interactions.
The simplest mathematical representation of this principle consists
on the third power of the adjacency matrix (A\textthreesuperior ).
Furthermore, Kovacs et al present a degree-normalized scaling for
this metric, which reduces bias caused by intermediate hubs within
the paths of length 3. This handcrafted measure, denoted L3, enables
the proposed approach to outperform previous methods for predicting
binary protein interactions in yeast (\emph{S. cerevisiae}), Arabidopsis
(\emph{A. thaliana}), worm (\emph{C. elegans}), fly (\emph{D. melanogaster}),
fission yeast (\emph{S. pombe}), mouse (\emph{M. musculus}) and humans
\cite{Kovacs2019}.

@@

The focus of this study is to evaluate different methods for predicting
PPIs using the network of known interactions. To achieve this, human
and rice PPI networks are compared using the proposed methods (A2,
A3, L3), as well as a low-dimensional representation of nodes (\texttt{Node2Vec})\cite{Grover_2016}.
After that, combinations of the two types of methods are tested. In
the case of the human network, two different versions of the human
interactome (\emph{HI-II-14}, \emph{HI-TESTED})\cite{Rolland2014Human}
are used for training and another interactome version is used for
validation (\emph{HI-III})\cite{Kovacs2019}. For the rice interactome,
a portion of the known interactions removed and then predicted with
different techniques based on the known interactions. The main contributions
of this paper are i) a general framework for link prediction in non-directed
networks and ii) two applications of this framework to biological
networks with a structural insight.

@@

Importance values for the A3 method stand out less from the node embedding
information than in A2 and L3. However, handcrafted features are more
important for the prediction of the human than in the rice interactome.

### 2020-12-09 nicolas/PPIPL3/ReportV7.tex
files were missing in git

\usepackage{babel}
\begin{document}
\title{Prediction of Protein-Protein Interactions on the Human and Rice Interactome }
\author{Nicolás López-Rozo\thanks{Pontificia Universidad Javeriana Cali, ORCID: 0000-0002-8060-6891\protect \\
\protect\hphantom{ema.}email: nicolaslopez@javerianacali.edu.co}, Jorge Finke\thanks{Pontificia Universidad Javeriana Cali, ORCID: 0000-0003-2393-6066},
Camilo Rocha\thanks{Pontificia Universidad Javeriana Cali, ORCID: 0000-0003-4356-7704}}
\maketitle
\begin{abstract}

Recent developments in network-based prediction of unmapped protein-protein
interactions (PPIs) reveal that two proteins are more likely to interact,
the higher the number of paths of length 3 between them, normalized
by the geometric average of their interactions. This paper extends
previous work on predicting binary interactions by taking into account
the learning of features (embeddings) of the PPI network. In particular,
we implement a gradient boosted decision tree model (\texttt{XGBoost})
using handcrafted features (including the normalized measure of paths
of length 3, L3) and embeddings from an algorithm that generates a
low-dimensional representation of nodes (\texttt{Node2Vec}). Our main
result shows that while L3 remains an important feature for predicting
PPIs, a better performance is achieved when L3 is combined with embedding
features. The proposed approach is validated for the human and rice
interactomes. For both cases, higher AUC values are obtained when
both types of features are used to train the model. The developed
framework can also be applied to interactomes of other organisms for
which a PPI network become recently available.
\end{abstract}

@@

A common way to create PPI networks or validate particular protein-protein
interactions is the \emph{Yeast-Two-Hybrid} technique (also known
as \emph{two-hybrid screening} or \emph{Y2H}). Figure \ref{Y2H}A
illustrates the biological basis of Y2H: the expression of a specific
reporter gene that activates by the binding of a DNA-binding Domain
(DB) and an Activation Domain (AD) of a Transcription Factor that
binds to an Upstream Activation Sequence (UAS). To evaluate an interaction,
the Y2H approach fuses a protein to the DB domain (known as \emph{bait})
and another protein to the AD (known as \emph{prey}). If the proteins
interact, then the reporter gene expression is activated by the AD
(Fig. \ref{Y2H}B). Otherwise, if proteins fail to interact, then
the reporter gene is not expressed (Fig. \ref{Y2H}C).

\begin{figure}[h]
\caption{\label{Y2H}The Yeast-2-Hybrid technique offers an experimental approach
for constructing PPI networks}

@@

The focus of this study is to evaluate different methods for predicting
PPIs using the network of known interactions. To achieve this, human
and rice PPI networks are compared using the proposed methods (A2,
A3, L3), as well as a low-dimensional representation of nodes (\texttt{Node2Vec})\cite{Grover_2016}.
After that, combinations of the two types of methods are tested. In
the case of the human network, two different versions of the human
interactome (\emph{HI-II-14}, \emph{HI-TESTED})\cite{Rolland2014Human}
are used for training and another interactome version is used for
validation (\emph{HI-III})\cite{Kovacs2019}. For the rice interactome,
a portion of the known interactions removed and then predicted with
different techniques based on the known interactions. The main contributions
of this paper are i) a general framework for link prediction in non-directed
networks and ii) two applications of this framework to biological
networks with a structural insight.

The remainder of this paper is organized as follows. Section 2 describes
the methodological steps and key milestones in the preparation of
the networks, model parameters and experimental configurations. Section
3 presents the main results of the models for the human and rice interactomes.
Section 4 presents the main conclusions of this study. Finally, Appendix
presents supplementary information and figures which complement the
results described in this paper. Figures in the Appendix can be identified
with a prefix (e.g. S\ref{F2}).

>>>>

\usepackage{babel}
\begin{document}
\title{Prediction of Protein-Protein Interactions on the Human and Rice Interactomes}
\author{Nicolás López-Rozo\thanks{Pontificia Universidad Javeriana Cali, ORCID: 0000-0002-8060-6891\protect \\
\protect\hphantom{ema.}email: nicolaslopez@javerianacali.edu.co}, Jorge Finke\thanks{Pontificia Universidad Javeriana Cali, ORCID: 0000-0003-2393-6066},
Camilo Rocha\thanks{Pontificia Universidad Javeriana Cali, ORCID: 0000-0003-4356-7704}}
\maketitle
\begin{abstract}

A recent study in network-based prediction of protein-protein interactions
(PPIs) reveals that two proteins are more likely to interact, the
higher the number of paths of length 3 between them (normalized by
the geometric average of their interactions). This paper extends previous
work on mapping binary interactions by taking into account the learning
of features (embeddings) of the PPI network. In particular, we implement
a gradient boosted decision tree model (\texttt{XGBoost}) using handcrafted
features (including the normalized measure) and embeddings from an
algorithm that generates a low-dimensional representation of nodes
(\texttt{Node2Vec}). Our main result shows that while the measure
remains an important feature for predicting interactions, better performance
is achieved when in addition embedding features are considered. The
proposed approach is validated for the human and rice interactomes.
For both cases, the combination of both types of features yield higher
AUC values. The developed framework can also be applied to interactomes
of other organisms for which PPI networks have recently become available.
\end{abstract}

@@

A common way to create PPI networks (or validate particular protein-protein
interactions) is the \emph{Yeast-Two-Hybrid} technique (also known
as \emph{two-hybrid screening} or \emph{Y2H}). Figure \ref{Y2H}A
illustrates the biological basis of Y2H: the expression of a specific
reporter gene is activated by the binding of a DNA-binding Domain
(DB) and an Activation Domain (AD) of a Transcription Factor, which
in turn binds to an Upstream Activation Sequence (UAS). To evaluate
an interaction between two proteins, the Y2H approach fuses one protein
to the DB domain (known as \emph{bait}) and another protein to the
AD (known as \emph{prey}). If the proteins interact, the reporter
gene expression is activated by the AD (Fig. \ref{Y2H}B). Otherwise,
if proteins fail to interact, the reporter gene is not expressed (Fig.
\ref{Y2H}C).

\begin{figure}[h]
\caption{\label{Y2H}The Yeast-2-Hybrid technique offers an experimental approach
for constructing PPI networks.}

@@

The focus of this study is to evaluate different methods for predicting
PPIs using the network of known interactions. To achieve this, human
and rice PPI networks are compared using the proposed methods (A2,
A3, L3), as well as a low-dimensional representation of nodes (\texttt{Node2Vec})\cite{Grover_2016}.
After that, combinations of the two types of methods are tested. In
the case of the human network, two different versions of the human
interactome \cite{Rolland2014Human} are used for training and another
interactome version is used for validation (\emph{HI-III})\cite{Kovacs2019}.
For the rice interactome, a portion of the known interactions removed
and then predicted with different techniques based on the known interactions.
The main contributions of this paper are i) a general framework for
link prediction in non-directed networks and ii) two applications
of this framework to biological networks with a structural insight.

The remainder of this paper is organized as follows. Section 2 describes
the methodological steps and key milestones in the preparation of
the networks, model parameters and experimental configurations. Section
3 presents the main results of the models for the human and rice interactomes.
Section 4 presents the main conclusions of this study. Finally, Appendix
presents supplementary information and figures which complement the
results described in this paper.
